From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/reversible1.tex
sec 2
Change and fix references
sec 2 and sec 3

The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic structure which we review in this section.

@@

In a reversible boolean circuit, the number of input bits matches the number of output bits. Circuits can be composed
sequentially, vertically, or horizontally, respectively preserving the number of bits, adding up the number of bits or
multiplying them. So, to program with reversible circuits, all we need to do is make sure that each primitive operation
preserves the number of bits, which is just a natural number. Natural numbers are the free commutative semiring (or,
commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication, so the commutative
semirig identities can be used to design a logic for reversible programming. This inspired the syntax for various
first-order reversible logics and reversible programming
languages~\cite*{sparksSuperstructuralReversibleLogic2014,jamesInformationEffects2012}.

Natural number identities have no computational content. To obtain a computational interpretation of the commutative rig
structure, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence~\cite{jamesInformationEffects2012}. This is a programming language which embodies the computational content
of isomorphisms of finite types, or permutations. On the semantic side, we need to consider the groupoidification of a
commutative rig, that is, a symmetric rig groupoid. A sound semantics for $\PiLang$ in weak rig groupoids was
established in~\cite{caretteComputingSemiringsWeak2016}, and conjectured to be complete.

% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

@@

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the left branch, which corresponds to
the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
type~\ensuremath{A}. The right branch which corresponds to the boolean being \textsf{false} passes the value of type $A$
through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the introduction
is transcribed as follows:

@@

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
(elided above) to re-shuffle the types. For the second circuit, instead of transcribing it directly, we express it using
a slightly more abstract notation:

@@

All of this is only half the story however as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions are collected in a second level of $\PiLang$
syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
level-1 type $A \isoone B$. For example, we have:

@@

%%%%%%%%%
\subsection{Denotational Semantics}

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be specified
by mapping each type to a finite set and each combinator to a permutation between finite sets. In this section, we
outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
combinators. This semantics will be extended in the next two sections to the HoTT metalanguage to properly model the
level-2 combinators.

The denotation of types is straightforward:

@@

\noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
is the cartesian product of sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping
$\denot{A}$ to $\denot{B}$:

@@

\begin{theorem}\label{thm:semone}
  The denotational semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
          combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
          extensional equivalence of functions.
  \end{itemize}
\end{theorem}
\begin{proof}
  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
    = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}

@@

Using categorical language, the denotational semantics for $\PiLang$ is using the category of finite sets and functions
$\SetFin$. However, we only use the bijective functions for the semantics, which means, we use the groupoid $\BFin =
  \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite
products $(\unit, \times)$, which we use to interpret our types. In $\BFin$, the coproducts and products restrict to
additive and multiplicative symmetric monoidal structures, respectively, making $\BFin$ a symmetric rig groupoid.

>>>>

The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic structure which we review in this section.

@@

In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides
sequential, vertical, and horizontal circuit composition operators.

These commutative semirig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence to obtain a computational interpretation of the commutative rig
structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
groupoidification of a commutative rig, that is, a symmetric rig groupoid.

Putting these ideas together, the programming language $\PiLang$, whose syntax is given below, embodies the
computational content of isomorphisms of finite types, or permutations.

% In programming parlance, as shown in~\Cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

@@

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the left branch, which corresponds to
the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
type~\ensuremath{A}. The right branch, which corresponds to the boolean being \textsf{false}, passes the value of type $A$
through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the previous section
is transcribed as follows:

@@

\addertwo{}

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second circuit, instead of transcribing it directly, we express it using
a slightly more abstract notation:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,Carette2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:

@@

%%%%%%%%%
\subsection{Semantics}
\label{subsec:denotational}

\noindent Below we present a semantic map that dually serves as an operational semantics and a denotational
semantics. As an operational semantics, each $\PiLang$ type $A$ is mapped to an Agda type $\denot{A}$ and each
combinator $c : A \isoone B$ is mapped to an Agda function $\denot{A} \to \denot{B}$. As a denotational semantics, each
$\PiLang$ type $A$ is mapped to a finite set $\denot{A}$ and each combinator $c : A \isoone B$ is mapped to bijection
between $\denot{A} \to \denot{B}$.

The semantics of types is straightforward:

@@

\noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
is the cartesian product of sets. The semantics of a combinator $c : A \isoone B$ is a (bijective) function mapping
$\denot{A}$ to $\denot{B}$:

@@

\begin{theorem}\label{thm:semone}
  The semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
          combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
          extensional equivalence of functions.
  \end{itemize}
\end{theorem}
\begin{proof}
  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
    = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}

@@

Using categorical language, the denotations of types and combinators are objects and morphisms in the category of finite sets and functions
$\SetFin$. However, as we only use the bijective functions for the semantics, we use the groupoid $\BFin =
  \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite
products $(\unit, \times)$, which we use to interpret our types. In $\BFin$, the coproducts and products restrict to
additive and multiplicative symmetric monoidal structures, respectively, making $\BFin$ a symmetric rig groupoid.
